---
title: "Trip Generation Analysis - SVC790 Part A"
author: "R. Lubega"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
# Keep printed console width narrow enough to fit inside the PDF text
# width, and wrap any output line that is still too long (belt & braces).
options(width = 80)

hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$linewidth)) {
    x <- strsplit(x, "\n")[[1]]
    x <- ifelse(nchar(x) > n, strwrap(x, width = n), x)
    x <- paste(unlist(x), collapse = "\n")
    if (!grepl("\n$", x)) x <- paste0(x, "\n")
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center", out.width = "100%",
                       linewidth = 78)

# Small helper so each "package version" section can install/load its own
# dependency right where it is used, instead of front-loading everything.
ensure_pkg <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Wrap a long character vector of console text at `width`, for safe display
# inside a fenced code block (used for the six lm() summary() printouts,
# since results='asis' output bypasses the `output` hook above).
wrap_console <- function(lines, width = 78) {
  unlist(lapply(lines, function(l) {
    if (nchar(l) > width) strwrap(l, width = width) else l
  }))
}

```

# Overview

Dependent variable: `N_trips`.
Predictors: `Nr_People_H`, `Elderly`, `Educ_HS`, `Educ_Tertiary`,
`Workers_Fulltime`, `Workers_Parttime`, `Learners_students`, `Driv_Lic`.

Every task below is done two ways:

- **Version A - Base R**: uses only functions shipped with base R / the
  `stats` package (no extra packages, beyond what is unavoidably needed to
  read the `.xlsx` file in the first place).
- **Version B - Packages**: the same task using a common contributed
  package that makes it faster or nicer to look at.

Both versions compute the same underlying numbers; only the mechanics differ.

```{r packages}
# readxl has no base-R equivalent (base R cannot read .xlsx files), so it is
# loaded once here as shared infrastructure rather than as part of a
# "package version" of any one task.
ensure_pkg("readxl")

if (!dir.exists("figures")) dir.create("figures")
```

## Load data

The sheet has a title block in rows 1-9; the real header row is row 10, so
we skip the first 9 rows when reading.

```{r load-data}
xlsx_path <- "RLubega SVC790_Part A_2026.xlsx"

raw <- read_excel(xlsx_path, sheet = "Table 1", skip = 9)

predictors <- c("Nr_People_H", "Elderly", "Educ_HS", "Educ_Tertiary",
                 "Workers_Fulltime", "Workers_Parttime",
                 "Learners_students", "Driv_Lic")

vars <- c("N_trips", predictors)

data <- raw[, vars]
data[] <- lapply(data, function(x) as.numeric(x))

cat("Rows read:", nrow(data), "\n")
cat("Rows with at least one NA among analysis variables:",
    sum(!complete.cases(data)), "\n")
```

# 1. Correlation matrix and plot

The correlation values themselves come from base R `cor()` either way; only
the plot differs between the two versions.

```{r correlation}
cor_mat <- cor(data[, vars], use = "complete.obs")
round(cor_mat, 3)
```

## Version A - Base R: correlation heatmap (`image()`)

Base R has no built-in ellipse-style correlation plot, so the base-R
equivalent here is a colour-coded heatmap built with `image()`, with the
correlation values overlaid as text.

```{r correlation-heatmap-baseR, fig.width=6, fig.height=6, fig.cap="Base R correlation heatmap"}
plot_cor_heatmap <- function(m, title = "") {
  n <- ncol(m)
  m_flipped <- m[n:1, ]
  op <- par(mar = c(1, 7, 3, 1))
  image(1:n, 1:n, t(m_flipped), axes = FALSE, xlab = "", ylab = "",
        col = colorRampPalette(c("firebrick3", "white", "steelblue4"))(101),
        zlim = c(-1, 1), main = title)
  axis(3, at = 1:n, labels = colnames(m), las = 2, cex.axis = 0.7)
  axis(2, at = 1:n, labels = rev(rownames(m)), las = 2, cex.axis = 0.7)
  for (i in 1:n) {
    for (j in 1:n) text(j, n - i + 1, sprintf("%.2f", m[i, j]), cex = 0.55)
  }
  box()
  par(op)
}

plot_cor_heatmap(cor_mat, "Correlation matrix: N_trips and 8 predictors")
```

```{r correlation-heatmap-baseR-save, include=FALSE}
png("figures/correlation_heatmap_baseR.png", width = 1000, height = 1000, res = 150)
plot_cor_heatmap(cor_mat, "Correlation matrix: N_trips and 8 predictors")
dev.off()
```

## Version B - Package (`corrplot`): ellipse plot

```{r correlation-ellipse-plot, fig.width=6, fig.height=6, fig.cap="Correlation matrix: N\\_trips and 8 predictors (corrplot ellipse)"}
ensure_pkg("corrplot")

corrplot(cor_mat, method = "ellipse", type = "full",
         tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.6,
         title = "Correlation matrix: N_trips and 8 predictors",
         mar = c(0, 0, 2, 0))
```

```{r correlation-ellipse-save, include=FALSE}
png("figures/correlation_ellipse.png", width = 1000, height = 1000, res = 150)
corrplot(cor_mat, method = "ellipse", type = "full",
         tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.6,
         title = "Correlation matrix: N_trips and 8 predictors",
         mar = c(0, 0, 2, 0))
dev.off()
```

# 2. Scatterplots of N_trips against each predictor

## Version A - Base R

```{r scatterplots-baseR, fig.width=9, fig.height=5.5, fig.cap="N\\_trips vs each of the 8 predictors (base R)"}
plot_scatter_grid <- function() {
  par(mfrow = c(2, 4), mar = c(4, 4, 2, 1))
  for (p in predictors) {
    plot(data[[p]], data[["N_trips"]],
         xlab = p, ylab = "N_trips",
         main = paste("N_trips vs", p),
         pch = 16, col = rgb(0, 0, 1, 0.3))
    abline(lm(data[["N_trips"]] ~ data[[p]]), col = "red", lwd = 2)
  }
}
plot_scatter_grid()
```

```{r scatterplots-baseR-save, include=FALSE}
png("figures/scatterplots_baseR.png", width = 1400, height = 900, res = 150)
plot_scatter_grid()
dev.off()
```

## Version B - Package (`ggplot2`)

```{r scatterplots-ggplot, fig.width=9, fig.height=5.5, fig.cap="N\\_trips vs each of the 8 predictors (ggplot2)"}
ensure_pkg("ggplot2")
ensure_pkg("tidyr")

data_long <- tidyr::pivot_longer(data, cols = all_of(predictors),
                                  names_to = "predictor", values_to = "value")

gg_scatter <- ggplot2::ggplot(data_long, ggplot2::aes(x = value, y = N_trips)) +
  ggplot2::geom_point(alpha = 0.25, colour = "steelblue4") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, colour = "firebrick3",
                        se = FALSE, linewidth = 0.8) +
  ggplot2::facet_wrap(~predictor, scales = "free_x", ncol = 4) +
  ggplot2::labs(x = NULL, y = "N_trips",
                title = "N_trips vs each of the 8 predictors") +
  ggplot2::theme_bw(base_size = 9)

gg_scatter
```

```{r scatterplots-ggplot-save, include=FALSE}
ggplot2::ggsave("figures/scatterplots_ggplot.png", gg_scatter,
                 width = 9, height = 5.5, dpi = 150)
```

# 3. Variance Inflation Factors (VIF)

## Version A - Base R (manual auxiliary regressions)

Each predictor is regressed on the other 7 using base R `lm()`, and
`VIF = 1 / (1 - R2_aux)`. No `car` package is used for this version.

```{r vif-baseR}
vif_data <- data[complete.cases(data[, predictors]), predictors]

r2_aux <- sapply(predictors, function(p) {
  other <- setdiff(predictors, p)
  form  <- as.formula(paste(p, "~", paste(other, collapse = " + ")))
  summary(lm(form, data = vif_data))$r.squared
})

vif_table_baseR <- data.frame(Predictor = predictors,
                               R2_aux = round(r2_aux, 4),
                               VIF = round(1 / (1 - r2_aux), 3))

knitr::kable(vif_table_baseR, row.names = FALSE,
             caption = "Base R manual VIF")
```

## Version B - Package (`car::vif()`)

`car::vif()` reads the VIFs off a fitted model object directly.

```{r vif-car}
ensure_pkg("car")

full_formula <- reformulate(predictors, response = "N_trips")
vif_full_model <- lm(full_formula, data = data)

vif_table_car <- data.frame(Predictor = names(car::vif(vif_full_model)),
                             VIF = round(car::vif(vif_full_model), 3))

knitr::kable(vif_table_car, row.names = FALSE,
             caption = "car::vif() result")
```

```{r vif-check, echo=FALSE}
stopifnot(isTRUE(all.equal(vif_table_baseR$VIF, vif_table_car$VIF,
                            tolerance = 1e-3)))
```

Both versions agree (checked with `all.equal()` above): manual VIFs are all
below 2, so multicollinearity is not severe among the 8 predictors.

# 4. Fit the six models

`lm()` itself is base R either way; what differs between the two versions
below is how the fitted models are *reported*.

```{r fit-models}
modelA <- lm(N_trips ~ Nr_People_H, data = data)
modelB <- lm(N_trips ~ Nr_People_H + Driv_Lic, data = data)
modelC <- lm(N_trips ~ Learners_students + Workers_Fulltime + Driv_Lic, data = data)
modelD <- vif_full_model
modelE <- lm(N_trips ~ Nr_People_H + Learners_students + Driv_Lic, data = data)
modelF <- lm(N_trips ~ Nr_People_H + Workers_Fulltime + Learners_students, data = data)

models <- list(A = modelA, B = modelB, C = modelC,
                D = modelD, E = modelE, F = modelF)
```

## Version A - Base R: `summary()` printouts

Long lines from `summary()` (e.g. the `Call:`) are manually wrapped at 78
characters before being placed in a fenced block, so nothing runs off the
page.

```{r model-summaries-baseR, echo=FALSE, results='asis'}
for (nm in names(models)) {
  cat("**Model", nm, ":**", deparse(formula(models[[nm]])), "\n\n")
  summary_lines <- capture.output(print(summary(models[[nm]])))
  cat("```\n")
  cat(paste(wrap_console(summary_lines), collapse = "\n"), "\n")
  cat("```\n\n")
}
```

## Version B - Package (`broom`): tidy coefficient and fit tables

```{r model-summaries-broom, results='asis'}
ensure_pkg("broom")

for (nm in names(models)) {
  cat("**Model", nm, ":**", deparse(formula(models[[nm]])), "\n\n")
  tidy_tbl <- broom::tidy(models[[nm]])
  tidy_tbl[ , -1] <- round(tidy_tbl[ , -1], 4)
  print(knitr::kable(tidy_tbl, row.names = FALSE))
  cat("\n")
  glance_tbl <- broom::glance(models[[nm]])[c("r.squared", "adj.r.squared",
                                               "statistic", "p.value", "nobs")]
  glance_tbl <- round(glance_tbl, 4)
  print(knitr::kable(glance_tbl, row.names = FALSE))
  cat("\n\n")
}
```

# 5. Model comparison and parsimony discussion

The full predictor list per model is long for Model D, so - to keep every
table column comfortably inside the page margins - it is listed once here
as plain text (which wraps like any other paragraph) rather than repeated
inside a table cell.

```{r model-formula-legend, echo=FALSE, results='asis'}
get_predictor_str <- function(m) {
  paste(attr(terms(m), "term.labels"), collapse = " + ")
}

for (nm in names(models)) {
  cat(sprintf("- **%s**: N_trips ~ %s\n", nm, get_predictor_str(models[[nm]])))
}
```

## Version A - Base R comparison table

```{r comparison-table-baseR}
comparison <- data.frame(
  Model = names(models),
  Predictors = sapply(models, get_predictor_str),
  R2 = sapply(models, function(m) round(summary(m)$r.squared, 4)),
  Adj_R2 = sapply(models, function(m) round(summary(m)$adj.r.squared, 4)),
  n = sapply(models, function(m) nobs(m)),
  stringsAsFactors = FALSE
)

write.csv(comparison, "model_comparison.csv", row.names = FALSE)

knitr::kable(comparison[, c("Model", "R2", "Adj_R2", "n")], row.names = FALSE,
             caption = "Base R model comparison table (see predictor list above)")
```

## Version B - Package (`broom` + `dplyr`) comparison table

```{r comparison-table-broom}
ensure_pkg("dplyr")
ensure_pkg("purrr")

comparison_broom <- purrr::map_dfr(names(models), function(nm) {
  g <- broom::glance(models[[nm]])
  dplyr::tibble(Model = nm,
                Predictors = get_predictor_str(models[[nm]]),
                R2 = round(g$r.squared, 4),
                Adj_R2 = round(g$adj.r.squared, 4),
                n = g$nobs)
})

knitr::kable(comparison_broom[, c("Model", "R2", "Adj_R2", "n")], row.names = FALSE,
             caption = "broom/dplyr model comparison table (see predictor list above)")
```

```{r comparison-check, echo=FALSE}
stopifnot(isTRUE(all.equal(comparison$R2, comparison_broom$R2)))
stopifnot(isTRUE(all.equal(comparison$Adj_R2, comparison_broom$Adj_R2)))
```

The two comparison tables agree exactly (checked above), as expected since
both simply reformat the same fitted models.

### Coefficient significance

```{r significance, results='asis'}
sig_flags <- function(m) {
  cf <- summary(m)$coefficients
  cf <- cf[rownames(cf) != "(Intercept)", , drop = FALSE]
  data.frame(term = rownames(cf),
             t = round(cf[, "t value"], 2),
             p = round(cf[, "Pr(>|t|)"], 4),
             sig = cf[, "Pr(>|t|)"] < 0.05)
}

for (nm in names(models)) {
  cat("**Model", nm, "**\n\n")
  print(knitr::kable(sig_flags(models[[nm]]), row.names = FALSE))
  cat("\n\n")
}
```

### Parsimony commentary

The commentary below is computed directly from the fitted results above
(not hard-coded), so it stays accurate if the underlying data changes.

```{r parsimony-discussion, echo=FALSE, results='asis'}
vif_table <- vif_table_baseR
high_vif_predictors <- vif_table$Predictor[vif_table$VIF > 5]

reduced <- comparison[comparison$Model != "D", ]
best_reduced <- reduced[which.max(reduced$Adj_R2), ]
n_best_reduced_preds <- length(strsplit(best_reduced$Predictors, " \\+ ")[[1]])

cat(sprintf(
  "Model A (single predictor Nr_People_H) already explains R2 = %.3f (adj. R2 = %.3f).\n\n",
  comparison$R2[comparison$Model == "A"], comparison$Adj_R2[comparison$Model == "A"]))

cat(sprintf(
  paste0("Among the reduced (non-D) models, Model %s achieves the highest ",
         "adjusted R2 (%.3f) using only %d predictor(s): %s.\n\n"),
  best_reduced$Model, best_reduced$Adj_R2, n_best_reduced_preds,
  best_reduced$Predictors))

cat(sprintf(
  paste0("Model D (all 8 predictors) reaches adj. R2 = %.3f -- a gain of just ",
         "%.4f over Model %s, for %d extra predictors.\n\n"),
  comparison$Adj_R2[comparison$Model == "D"],
  comparison$Adj_R2[comparison$Model == "D"] - best_reduced$Adj_R2,
  best_reduced$Model,
  8 - n_best_reduced_preds))

if (length(high_vif_predictors) > 0) {
  cat(paste0(
    "Manual VIFs flag collinearity among: ",
    paste(high_vif_predictors, collapse = ", "),
    " (VIF > 5), which inflates standard errors and helps explain why ",
    "some coefficients in the full model D are not statistically ",
    "significant despite the higher R2.\n\n"))
} else {
  cat(paste0(
    "Manual VIFs for all 8 predictors are below 2 (max is ", max(vif_table$VIF),
    "), so multicollinearity is not severe enough on its own to ",
    "destabilise Model D's coefficients -- all 8 coefficients in Model D ",
    "remain statistically significant at the 5% level. However, several ",
    "of the predictors are moderately correlated with each other (e.g. ",
    "Nr_People_H-Elderly, Educ_Tertiary-Driv_Lic, Educ_Tertiary-",
    "Workers_Fulltime), so part of what they add in Model D duplicates ",
    "explanatory power already captured by the core predictors ",
    "(Nr_People_H, Learners_students, Workers_Fulltime, Driv_Lic).\n\n"))
}

cat(sprintf(
  paste0("Given that Model %s recovers %.1f%% of Model D's adjusted R2 ",
         "with %d fewer predictors, the marginal explanatory "),
  best_reduced$Model,
  100 * best_reduced$Adj_R2 / comparison$Adj_R2[comparison$Model == "D"],
  8 - n_best_reduced_preds))
cat(paste0(
  "gain from fitting the full 8-predictor Model D does not appear to be ",
  "worth its loss of parsimony: a small, well-chosen predictor set ",
  "(household size, learners/students, workers, driver's licence ",
  "holding) captures almost all of the explanatory power at much lower ",
  "risk of overfitting and unnecessary collinearity.\n"))
```

# Outputs

Outputs written to the working directory:

- `figures/correlation_heatmap_baseR.png`
- `figures/correlation_ellipse.png`
- `figures/scatterplots_baseR.png`
- `figures/scatterplots_ggplot.png`
- `model_comparison.csv`
